---
title: "Correlation Analysis using Monte Carlo Hypothesis Test"
author: "Jingyan Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(usmap)
```

# Prepare the original data

```{r}
shooting = read.csv("shooting.csv")
code = read.csv("codebook.csv")
shooting = na.omit(shooting)
shooting = select(shooting, -Sources)
colnames(shooting) <- tolower(gsub("[^[:alnum:]_]", " ", colnames(shooting)))
colnames(shooting) <- gsub("[.]", "_", colnames(shooting))
shooting$date <- as.character(shooting$date)
modify_date <- function(date) {
  parts <- strsplit(date, "/")[[1]]
  if (length(parts) == 3) {
    if (parts[2] == "Unknown") {
      return(paste(parts[1], "XX", parts[3], sep="/"))
    } else {
      return(date)
    }
  } else {
    return(date)
  }
}

shooting$date <- sapply(shooting$date, modify_date)
cols_to_factor <- names(shooting)[sapply(shooting, is.character)]
shooting[cols_to_factor] <- lapply(shooting[cols_to_factor], as.factor)
```


# Use Monte Carlo Hypothesis Test on Generated Different Distribution


## Function: Linearity Test
Test the linearity between two data sets, return the p-value for linearity. 
```{r}
# Perform Monte Carlo hypothesis test to assess linearity
# Define null hypothesis: No linear relationship
# Alternative hypothesis: Linear relationship

linear_monte_carlo <- function(x,y){

n_simulations <- 1000
data_df <- data.frame(x=x,y=y)
# Initialize a vector to store simulated correlation coefficients
simulated_correlations <- numeric(n_simulations)

# Observed correlation coefficient
observed_correlation <- cor(data_df$x, data_df$y, method = "pearson")

# Perform Monte Carlo simulations
for (i in 1:n_simulations) {
  shuffled_data <- data_df %>%
    mutate(y = sample(y))
  
  # Calculate the correlation coefficient for the shuffled data
  simulated_correlation <- cor(shuffled_data$x, shuffled_data$y, method = "pearson")
  
  # Store the simulated correlation coefficient
  simulated_correlations[i] <- simulated_correlation
}

# Calculate the two-tailed p-value
p_value <- mean(abs(simulated_correlations) >= abs(observed_correlation))
return(p_value)
}
```

## Function: Plot the p-value

```{r}
plot_p <- function(results){
  plot(1:nrow(results), results$p_value, type = "l", ylab = "P-value", xlab = "Simulation Run")
  abline(h = 0.05, col = "red", lty = 2)
}
```



## Normal Distribution v.s. Normal Distribution
```{r}
# Set a random seed for reproducibility
set.seed(123)

# Generate two datasets
k = 100
Group_A <- rnorm(k, mean = 0, sd = 1)
means = seq(from = -1, to = 1, length.out = 10)
sds = seq(from = 0.1, to = 2, length.out = 10)
results <- expand.grid(mean = means, sd = sds)
results$p_value <- NA # Initialize a column for p-values

for (i in 1:nrow(results)) {
    Group_B <- rnorm(k, mean = results$mean[i], sd = results$sd[i])
    results$p_value[i] <- linear_monte_carlo(Group_A, Group_B)
}
plot_p(results)
```

```{r}
results[results$p_value<=0.05,]
```



## Normal Distribution v.s. Skewed Distribution


### Right-skewed 
```{r}
set.seed(123)
for (i in 1:nrow(results)) {
    Group_B <- rlnorm(k, mean = results$mean[i], sd = results$sd[i])
    results$p_value[i] <- linear_monte_carlo(Group_A, Group_B)
}
plot_p(results)
```
```{r}
results[results$p_value<=0.05,]
```


 
## Normal Distribution v.s. Exponential Distribution

```{r}
set.seed(123)
Group_A <- rnorm(k, mean = 0, sd = 1)
lambda = seq(from = 0.1, to = 2, length.out = 100)
results <- expand.grid(lambda = lambda)
results$p_value <- NA # Initialize a column for p-values

for (i in 1:nrow(results)) {
    Group_B <- rexp(k, rate = lambda)
    results$p_value[i] <- linear_monte_carlo(Group_A, Group_B)
}
plot_p(results)
```
```{r}
results[results$p_value<=0.05,]
```

 
## Uniform Distribution v.s. Uniform Distribution

```{r}
set.seed(123)
Group_A <- runif(k, min = 3, max = 7)

mins = seq(from = 1, to = 5, length.out = 10)
maxs = seq(from = 5, to = 10, length.out = 10)
results <- expand.grid(min = mins, max = maxs)
results$p_value <- NA # Initialize a column for p-values

for (i in 1:nrow(results)) {
    Group_B <- runif(k, results$min[i], results$max[i])
    if (sd(Group_B) == 0) {
        # Handle the case where the standard deviation is zero (e.g., skip or assign NA)
        results$p_value[i] <- NA
    } else {
        results$p_value[i] <- linear_monte_carlo(Group_A, Group_B)
    }
}
plot_p(results)
```
```{r}
results[results$p_value<=0.05,]
```


The method works fine when comparing uniform distribution.



# Analysis

```{r}
results_df <- data.frame(
  Predictor = character(0),  
  P_Value = numeric(0)       
)

for (i in 1:ncol(num_shooting)) {
  predictor_name <- colnames(num_shooting)[i]  # Get the predictor's name
  p_value <- linear_monte_carlo(shooting$`total injured killed victims`, num_shooting[, i])
  results_df <- rbind(results_df, data.frame(Predictor = predictor_name, P_Value = p_value))
}

print(results_df)
```

```{r}
cor(shooting$`shooter age`, shooting$`total injured killed victims`)
```


```{r}
plot(shooting$`reliability score  1 5 `, shooting$`total injured killed victims`)
abline(lm(`total injured killed victims`~`reliability score  1 5 `, data = shooting), col = "red")
```


```{r}
plot(shooting$`shooter age`, shooting$`total injured killed victims`)
abline(lm(`total injured killed victims`~`shooter age`, data = shooting), col = "red")
```


